---
title: 'Introduction to R^[`r library(r2symbols) ; format(paste(symbol("copyright") , " - Wim R.M. Cardoen, 2022 - The content can neither be copied nor distributed without the **explicit** permission of the author."))`]'
subtitle: 'Section 4: Control-flow and functions'
author: "Wim R.M. Cardoen"
date: "Last updated: `r format(Sys.time(), ' %2m/%2d/%Y @ %2H:%2M:%2S')`"
output:
  html_document:
    highlight: tango
    df_print: tibble
    toc: true
    toc_depth: 5
    number_sections: true
    mathjax: default
bibliography: [latex/intro.bib]
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/research-institute-for-nature-and-forest.csl
lang: en
---

# R Control flow

*"Stay out of the loop, the club, the inner circle." — Michael Leunig*

## Conditional constructs

* `if`, `else if`, `else`

  Syntax:
  ```{R, echo=TRUE, eval=FALSE}
  if(condition1){
      ...
  }[else if(condition2){
      ...
  }else{
      ...
  }]
  ```

* `ifelse()`

  The R language provides a **vectorized** version: `ifelse()`.

  **Syntax**:
  ```{R, echo=TRUE, eval=FALSE}
  ifelse(condition, vecy, vecn)
  ```
  where:
  - *condition*: test condition
  - *vecy*: values in case of `TRUE` values
  - *vecn*: values in case of `FALSE` values

  There exists a more general version of `ifelse()` i.e.
  [dplyr::case_when()](https://dplyr.tidyverse.org/reference/case_when.html)


### Examples
```{R, echo=TRUE, comment=''}
score <- 75.0

if(score>=90.0){
  grade <- 'A'
} else if((score<90.0) && (score>=80.0)){
  grade <- 'B'
} else if((score<80.0) && (score>=70.0)){
  grade <- 'C'
} else{
  grade <- 'D'
}
cat(sprintf("Score:%4.2f -> Grade:%s\n", score, grade))
```

```{R, echo=TRUE, comment=''}
x <- c(-1,2,1,-5,-7)
c
res <- ifelse(x>=0, x,-x)
res
```

### Exercises

* Write code to find the smallest of three numbers, e.g. 21, 12, 17 using
  an `if`, (`else if`, `else`) construct.

## Loop constructs

There are several loop constructs:

* `while`

  ```{R, echo=TRUE, eval=FALSE}
  while(condition){
    body of the loop 
  }
  ```

* `for`

  ```{R, echo=TRUE, eval=FALSE}
  for(item in sequence){
    body of the loop 
  }
  ```

* `repeat`

  ```{R, echo=TRUE, eval=FALSE}
  repeat{
    body of the loop 
  }
  ```
  The `repeat` loop has no condition to leave the loop: insert a `break`.

The `break` statement allows one to break out of the `while`, `for` and `repeat` constructs.
The `next` statement allows one to go to the next iteration.

### Examples

* for loop construct

```{R, echo=TRUE, comment=''}
# Loop over all items
fruit <- c("apple", "pear", "banana", "grape")
for(item in fruit){
  cat(sprintf(" Fruit:%s\n", item))
}
```

```{R, echo=TRUE, comment=''}
# Skip all numbers which are multiples of 3
x <- sample(1:100, size=10, replace=FALSE)
x
```

```{R, echo=TRUE, comment=''}
for(item in x){
  if(item%%3==0)
     next
  cat(sprintf(" %3d is NOT a multiple of 3\n", item))
}
```

* while loop

```{R, echo=TRUE, comment=''}
x <- sample(1:1000, size=100, replace= FALSE)
isFound <- FALSE
i <- 1
while(!isFound){
  if(x[i]%%7==0){
      cat(sprintf(" %3d is divisible by 7\n", x[i]))
      isFound <- TRUE
  }
  else{
      cat(sprintf(" %3d is NOT divisible by 7\n", x[i]))
      i <- i + 1
  }
}
```

* repeat loop

```{R, echo=TRUE, comment=''}
i <- 1
repeat{
  # Stop the loop as soon as you find a multiple of 7.
  if(x[i]%%7==0){
      cat(sprintf(" %3d is divisible by 7\n", x[i]))
      break
  }
  else{
      cat(sprintf(" %3d is NOT divisible by 7\n", x[i]))
      i <- i + 1
  }
}
```

### Exercises

* The [**Fibonacci** sequence](https://en.wikipedia.org/wiki/Fibonacci_number) is defined by the following recurrence relation:

  $$F_n = F_{n-1} + F_{n-2} \;, \quad \forall \; n \in \mathbb{N}$$

  where $F_1 = F_2 = 1$.
  Calculate all Fibonacci numbers up to $F_{15}$, i.e.:

  ```{R, echo=FALSE, eval=TRUE, comment=''}
  x <- vector(mode="integer", length=15)
  x[1] <- 1
  x[2] <- 1
  for(i in 3:15){
      x[i] <- x[i-1] + x[i-2]
  }
  x
  ```

* The square root of a number $n$ is equivalent to solving the following equation:

  $$x^2 - n = 0$$

  The solution to this equation can be found iteratively by using e.g. the
  [Newton-Raphson method](https://en.wikipedia.org/wiki/Newton%27s_method).

  Iteration $i+1$ for $x$ is then given by:

  $$x_{i+1} = \frac{1}{2}\left(x_i + \frac{n}{x_i}\right)$$

  Find the square root of $751$ to a precision of at least 8 decimals. You can set $x_0$ to $n$ itself.

---

# R Functions

## General statements

* The **most common way** to create a function is to assign a function name **and**
  use the `function()` statement.

  **Syntax**:

  ```{R, echo=TRUE, eval=FALSE}
  function_name <- function(arg_list){
      # body of the function
  }
  ```

  An **anonymous** function does **not** bear a name. It can be useful for short expressions.

* A function in R has 3 important components:
  * arguments of the function: `formals()`
    R allows default arguments. The default values are assigned using the `=` sign.
  * body of the function: `body()`
  * environment in which the function runs: `environment()`

  Primitive functions (e.g. `prod()`) call directly C code with `.Primitive()` (no R code involved).
  These primitive functions can be found in `package:base`.
  If the functions `formals()`, `body()` and `environment()`
  are applied to primitive functions `NULL` will be returned.

* A function can exit in 2 ways:
  * by returning a value:
    * implicit return: last expression evaluated in the body
    * explicit return: by invoking the `return()` function
  * through error e.g. by invoking the `stop()` function

  An R function returns **only** 1 object.
  If you want to return more than 1 object, put the objects in an R `list()`
  and return the list^[The R concept of a list will be discussed in a later section.].

* Assignment of variables in functions:

  Ordinary assignments performed within the function are local and temporary to the function.
  After the function has been exited the values of these assignments are lost.

  If global and permanent assignment are needed within a function, you can use the
  superassignment operator `<<-` or the `assign()` function.

* Pass by value or pass by reference:

  By default R does a **pass by value**. The **pass by reference** is possible in R but
  requires the use/understanding of the concept of an R environment^[The R concept of an environment will be discussed in a later section.].


### Examples

* Implicit return
```{R, echo=TRUE, comment=''}
a <- 2.0
b <- 8.0
myprod1 <- function(x,y){
  x*y
}
cat(sprintf(" Prod of %.1f and %.1f is %.1f\n", a, b, myprod1(a,b)))
```

* Use of an explicit return statement
```{R, echo=TRUE, comment=''}
myprod2 <- function(x,y){
  return(x*y)
}
cat(sprintf(" Prod of %.1f and %.1f is %.1f\n", a, b, myprod2(a,b)))
```

* Default argument
```{R, echo=TRUE, comment=''}
myshift <- function(x, shift=1){
  return(x+shift)
}
x <- seq(from=2, to=100, length=5)
x
y1 <- myshift(x)
y1

y3 <- myshift(x,3)
y3
```

* Anonymous functions (i.e. function without a name)
```{R, echo=TRUE, comment=''}
# sapply: apply a function on a vector.
inp <- 1:10
inp
out <- sapply(inp, function(x){return(x%%2==0)})
out
```

  Numerical integration of $\displaystyle \int_{0}^{1} \frac{4}{1+x^2} \, dx$

```{R, echo=TRUE, comment=''}
res <- integrate(function(x){4.0/(1.0+x^2)}, 0.0, 1.0)
res
```

* Retrieve the formal arguments of a function
```{R, echo=TRUE, comment=''}
formals(myprod2)
```

* Retrieve the body of a function
```{R, echo=TRUE, comment=''}
body(myprod2)
```

* Retrieve the environment of a function
```{R, echo=TRUE, comment=''}
environment(myprod2)
```

* Local vs. global variables

```{R, echo=TRUE, comment=''}
x <- 1     # x: in Global Environment 
myfunc1 <- function(){
  x <-5
}
res <- myfunc1()
cat(sprintf(" Value of x:%d\n",x))
```

```{R, echo=TRUE, comment=''}
myfunc2 <- function(){
  x <<-5
}
res <- myfunc2()
cat(sprintf(" Value of x:%d\n",x))
```

* Pass by value
```{R, echo=TRUE, comment=''}
x <- 1:5
cat(sprintf("BEFORE call:\n"))
x

mychange <- function(x)
{
   x <- 5:1
}
res <- mychange()
cat(sprintf("AFTER call:\n"))
x
```

### Exercises

* Write your own factorial function named `myfactorial(n)`.
  The factorial function, $n!$ is defined as:

  $$n! = n\,(n-1)!$$

  where $0! := 1$.

* Write your own function named `castdie(n)` which simulates casting a die $n$ times.
  - Assume you have a fair die.
  - Adjust the function `castdie(n)` for the general case i.e. a non-fair die,
    e.g. $p=\left(\frac{1}{10},\frac{1}{10},\frac{3}{10},\frac{3}{10},\frac{1}{10},\frac{1}{10}\right)$
  - Hint: use R's `sample()` function.

* An auto-regressive time series of type AR(1) is defined as follows (see e.g. [@BROCKWELL:2009a]):

  $$x_i = \varphi \, x_{i-1}\, + \, \varepsilon_i$$

  where $\varepsilon_i \, \sim \, N(0,\sigma^2)$ and are iid^[iid: identically and independently distributed.].

  - Write the function `genAR1Series(n=1000, x0=0.0, phi=0.7)`
    which returns the AR(1) time series $\{x_i\}$ for $i \in \{1, \ldots, n\}$, where:

    $$x_0 = 0.0, \quad \varphi = 0.7, \quad \varepsilon_i \sim N(0,1) \;, \; \forall \, i \, \in \, \{1, \ldots, n\}$$

  - Write the sample autocorrelation function (`myacf(x)`) (ACF)
    which calculates a vector of $\rho(h)$'s
    where the lag $h \, \in \, \{0, 1, 2, \ldots, n-1\}$.

    The autocorrelation with lag $h$, i.e. $\rho(h)$ is defined as follows:

    $$\rho(h) := \frac{\gamma(h)}{\gamma(0)}$$

    where:

    $$\gamma(h) := \frac{1}{n} \sum_{i=1}^{n-h} \Big( x_{i+h} \,-\,\overline{x} \Big)\Big( x_i \,-\,\overline{x} \Big)$$

    $$\overline{x} := \frac{1}{n} \sum_{i=1}^n x_i$$

  - Calculate the autocorrelation vector for the time series you generated previously.
    You can check your results with R's `acf()`
    function^[For the AR(1) model, the theoretical value of $\rho(h)$ is given by $\varphi^h$.].

## Prefix vs. infix functions

* most functions are prefix: the name of the functions **precedes** the arguments.
  ```{R, echo=TRUE, comment=''}
  res <- sum(1,2)
  res
  ```

* infix: the function name/operator is found between the arguments
  ```{R, echo=TRUE, comment=''}
  res <- 1 + 2
  res
  ```

  In R you can create your own infix operator by defining a function as follows:
  ```{R, echo=TRUE, eval=FALSE}
  '%op%' <- function(x,y){
      # Body of the function
  }
  ```

  The infix function is then invoked as follows:
  ```{R, echo=TRUE, eval=FALSE}
  x %op% y
  ```

* R contains some **predefined** infix operators:
  * `%%`: modulo operator
  * `%/%`: integer division
  * `%*%`: matrix multiplication
  * `%o%`: outer product
  * `%x%`: Kronecker product
  * `%in%`: matching operator


### Example

* Let $\mathbf{x}, \mathbf{y} \, \in \, \mathbb{R}^n$. The angle $\theta$ between $\mathbf{x}$ and $\mathbf{y}$ is given by:

  $$\theta = \arccos{\left( \frac{\mathbf{x} \cdot \mathbf{y}}{\|\mathbf{x}\|\,\|\mathbf{y}\|} \right)}$$

  ```{r, echo=TRUE, comment=''}
  '%theta%' <- function(x,y){
      nom <- sum(x*y)
      denom <- sqrt(sum(x^2)) *sqrt(sum(y^2))
      return(acos(nom/denom))
  }
  ```

  ```{r, echo=TRUE, comment=''}
  x <- c(1,0,3)
  x
  y <- c(3,2,-1)
  y
  cat(sprintf("The angle (radians) between x and y is:%8.4f\n", x %theta% y))
  ```

### Exercises

* In the cyclic group $\mathbb{Z}_4$, we only have the (integer) elements: $\{0,1,2,3\}$.
  The addition ($\forall\,x,y \, \in \,\mathbb{Z}_4$) is defined as follows:

  $$x+y \, \equiv \, x+y \mod 4$$

  ```{R, echo=FALSE}
  '%+%' <- function(x,y){
     return((x+y)%% 4)
  }
  ```
  ```{R, echo=TRUE, comment=''}
  x <- sample(0:3, size=10, replace=TRUE)
  x
  y <- sample(0:3, size=12, replace=TRUE)
  y
  ```
  Invoking the infix addition `%+%` results into:
  ```{R, echo=TRUE, comment=''}
  res <- x %+% y
  res
  ```
  Write the infix function (`%+%`) to perform addition in the cyclic group $\mathbb{Z}_4$.

## Function composition/piping

Often the output of a function is used as the input for another function.
There are several options beyond the explicit creation of intermediate variables:

* Function composition (e.g. `h(g(f(x)))`).

  ```{R, echo=TRUE, comment=''}
  x <- runif(1000)
  xav <- mean(x)
  res <- sqrt(sum((x-xav)^2))  
  res
  ```

* Use of piping (cfr. the following Linux code: `cat *.r | grep function`).
  The [magrittr library](https://magrittr.tidyverse.org/) was developed to
  provide this feature^[The infix function `%>%` is known as the "and then" operator.] among others.

  ```{R, echo=TRUE, comment=''}
  library(magrittr)
  (x-xav)^2 %>% 
      sum() %>% 
      sqrt()
  ```

# Bibliography
